# Data

## Description
The data sources we will be using for this project are: 
1) Public housing: HDB Resale Flat Prices dataset (https://data.gov.sg/datasets?agencies=HDB&resultId=d_b51323a474ba789fb4cc3db58a3116d4)
2) Private housing: Private Residential Property Transactions
(https://www.ura.gov.sg/Corporate/Property/Property-Data/Private-Residential-Properties)

Both datasets are provided by official government ministries of Singapore, with the first being from the Housing and Development Board (HDB) and the second from the Urban Redevelopment Authority (URA). Both datasets are updated quarterly, and we will focus on data from the past 5 years to capture recent trends and the impact of government cooling measures.

## Missing value analysis
At first glance, the dataset appears to have NA and “-” to indicate missing values. This likely represents represents "Not Applicable/Not Available" and "No Transactions" respectively. The NA value likely means that these type of flats are not available for purchase in those neighbourhoods (Bishan & 1-Room Flats) and the "-" could mean that there are flats available but no purchases were made. We will load the data and check for missing values across all columns.

```{r}
# Load necessary libraries
library(tidyverse)
library(sf)
library(lubridate)

# 1. Load the Data
df <- read_csv("datasets/MedianResalePricesforRegisteredApplicationsbyTownandFlatType.csv", 
               na = c("na", "-", "NA"))

# Load Singapore Geospatial Data (Planning Areas)
# We use a public GeoJSON link for Singapore Planning Areas
# If this link fails, you can download "Master Plan 2019 Planning Area Boundary" from data.gov.sg
geojson_url <- "https://raw.githubusercontent.com/yinshanyang/singapore/master/maps/2-planning-area.geojson"
sg_map <- st_read(geojson_url, quiet = TRUE) %>%
  mutate(name = toupper(name)) # Standardize map names to Uppercase

# 2. Check for Missing Values
missing_summary <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "missing_{col}"))
print(missing_summary)

# 3. Clean Dataset
df_clean <- df %>%
  mutate(
    # Convert 'price' to numeric (this handles any remaining non-numeric values)
    price = as.numeric(price),
    # Convert 'quarter' (e.g., "2007-Q2") to a Date object
    date = yq(quarter),
    # Standardize town names to uppercase
    town = toupper(town)
  ) %>%
  # Remove rows where price is missing
  filter(!is.na(price))
# 4. Preview Cleaned Data
print(head(df_clean))
```

## Plotting Dataset 
We will create some initial plots to visualize the data and identify trends or patterns.
```{r}
# Plot A: Trend of Average Median Resale Prices by Flat Type 
# Compute average price per flat type over time
avg_price_trend <- df_clean %>%
  group_by(date, flat_type) %>%
  summarise(avg_price = mean(price, na.rm = TRUE)) %>%
  ungroup()
# Plot the trend
ggplot(avg_price_trend, aes(x = date, y = avg_price, color = flat_type)) +
  geom_line() +
  labs(title = "Trend of Median Resale Prices by Flat Type",
       x = "Date",
       y = "Average Price (SGD)",
       color = "Flat Type") +
  theme_minimal()
```

```{r}
#| fig-width: 20
#| fig-height: 15
#| 
# Plot B: Trend of 4-Room Resale Flat Prices in Selected Towns by Region
# 1. Define Towns and Regions 
north_towns <- c("WOODLANDS", "YISHUN", "SEMBAWANG")
northeast_towns <- c("SENGKANG", "HOUGANG", "PUNGGOL")
east_towns  <- c("TAMPINES", "BEDOK", "PASIR RIS")
west_towns  <- c("JURONG WEST", "CHOA CHU KANG", "BUKIT BATOK")
central_towns <- c("BISHAN", "TOA PAYOH", "GEYLANG")
selected_towns <- c(north_towns, northeast_towns, east_towns, west_towns, central_towns)

# Map towns to regions
region_map <- data.frame(
  town = selected_towns,
  region = c(rep("North", 3), rep("North-East", 3), rep("East", 3), rep("West", 3), rep("Central", 3))
)

# 2. Filter Data
df_4room_selected <- df_clean %>%
  filter(flat_type == "4-room") %>%
  filter(town %in% selected_towns) %>%
  left_join(region_map, by = "town")

# 3. Create Label Data (New Step)
# We calculate the last data point for each town to know where to put the label
town_labels <- df_4room_selected %>%
  group_by(town) %>%
  filter(date == max(date)) %>%
  ungroup()

# 4. Plotting
ggplot(df_4room_selected, aes(x = date, y = price, color = town)) +
  geom_line(size = 1) +
  # Add labels directly to the end of the lines
  geom_text(data = town_labels, aes(label = town), 
            hjust = -0.1, size = 3, show.legend = FALSE) +
  # Use facet_wrap without 'scales = "free_y"' to keep Y-axis common
  facet_wrap(~region) + 
  # Expand the x-axis slightly to the right to make room for the text labels
  scale_x_date(expand = expansion(mult = c(0.05, 0.4))) +
  labs(
    title = "4-Room Resale Flat Prices by Region (15 Selected Towns)",
    subtitle = "Quarterly Median Resale Prices (Common Y-Axis)",
    x = "Year",
    y = "Price (SGD)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none" # Remove the legend since lines are labeled directly
  ) +
  scale_y_continuous(labels = scales::comma)
```

```{r}
# Plot C: Choropleth Map of Average 4-Room Resale Flat Prices by Planning Area for Target Years
# Process & Filter Price Data
# We need to map HDB town names to URA Planning Area names
target_years <- c(2010, 2015, 2020, 2025)

df_map_data <- df %>%
  mutate(
    price = as.numeric(price),
    date = yq(quarter),
    year = year(date),
    town = toupper(town)
  ) %>%
  filter(flat_type == "4-room") %>%
  filter(year %in% target_years) %>%
  filter(!is.na(price)) %>%
  # --- Crucial Step: Fix Name Mismatches ---
  mutate(town = case_when(
    town == "KALLANG/WHAMPOA" ~ "KALLANG",   # Map to closest major planning area
    town == "CENTRAL AREA"    ~ "DOWNTOWN CORE",
    TRUE ~ town
  )) %>%
  # Calculate average median price for the whole year
  group_by(year, town) %>%
  summarise(avg_price = mean(price, na.rm = TRUE)) %>%
  ungroup()

# 4. Prepare Geospatial Data for Faceting
# To facet correctly (keeping empty regions grey), we create a grid of All Regions x All Years
full_grid <- expand_grid(
  name = unique(sg_map$name),
  year = target_years
)

# Join the map shapes with the grid, then join the price data
plot_data <- full_grid %>%
  left_join(sg_map, by = "name") %>%
  st_as_sf() %>% # Convert back to sf object after join
  left_join(df_map_data, by = c("name" = "town", "year"))

# 5. Plot the Faceted Choropleth Map
ggplot(plot_data) +
  geom_sf(aes(fill = avg_price), color = "white", size = 0.2) +
  facet_wrap(~year, ncol = 2) +
  # Use a consistent scale across all facets
  scale_fill_viridis_c(
    option = "magma", 
    direction = -1,  # Darker colors = Higher prices
    na.value = "grey90", 
    labels = scales::comma,
    name = "Avg Price (SGD)"
  ) +
  labs(
    title = "Evolution of 4-Room Resale Flat Prices (2010 - 2025)",
    subtitle = "Average Median Resale Price by Town",
    caption = "Grey areas indicate no HDB resale data available for that year"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(size = 12, face = "bold")
  )
```
```{r}
# Lease Decay: Panel Plot + GAM + Heatmap --------------------------

library(tidyverse)
library(lubridate)
library(stringr)
library(mgcv)

# 1. Load transaction-level HDB resale data (2017 onwards)
hdb_resale <- read_csv(
  "datasets/ResaleflatpricesbasedonregistrationdatefromJan2017onwards.csv",
  show_col_types = FALSE
)

# 2. Clean remaining lease + compute price per sqm -----------------

hdb_resale_clean <- hdb_resale %>%
  mutate(
    years  = str_extract(remaining_lease, "\\d+(?= year)")  %>% as.numeric(),
    months = str_extract(remaining_lease, "\\d+(?= month)") %>% as.numeric(),
    months = replace_na(months, 0),
    remaining_lease_years = years + months / 12,
    price_psm = resale_price / floor_area_sqm
  ) %>%
  filter(
    remaining_lease_years >= 0,
    remaining_lease_years <= 99,
    price_psm > 0
  )

main_flats <- c("3 ROOM", "4 ROOM", "5 ROOM")

panel_data <- hdb_resale_clean %>%
  filter(flat_type %in% main_flats)

# 3. Panel plot: price per sqm vs remaining lease, by flat type ----

ggplot(panel_data, aes(x = remaining_lease_years, y = price_psm)) +
  geom_point(alpha = 0.05, size = 0.6) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, color = "darkred") +
  facet_wrap(~ flat_type, scales = "free_y") +
  labs(
    title = "Lease Decay in HDB Resale Prices (2017–Present)",
    subtitle = "Price per sqm vs remaining lease, by flat type",
    x = "Remaining Lease (years)",
    y = "Resale Price per sqm (SGD)"
  ) +
  theme_minimal()

# 4. GAM curve: smooth lease-decay shape ---------------------------

gam_data <- panel_data   # reuse

gam_model <- gam(
  price_psm ~ s(remaining_lease_years) + flat_type,
  data = gam_data
)

pred_grid <- expand_grid(
  remaining_lease_years = seq(0, 99, by = 0.5),
  flat_type = main_flats
) %>%
  mutate(pred = predict(gam_model, newdata = cur_data_all()))

ggplot(pred_grid, aes(x = remaining_lease_years, y = pred, color = flat_type)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Smooth GAM Curve of Lease Decay (HDB Resale Market)",
    subtitle = "Predicted price per sqm by remaining lease and flat type",
    x = "Remaining Lease (years)",
    y = "Predicted Price per sqm (SGD)",
    color = "Flat Type"
  ) +
  theme_minimal()

# 5. Heatmap: lease bins × flat type -------------------------------

hdb_lease_decay <- panel_data %>%
  mutate(
    lease_bin = cut(
      remaining_lease_years,
      breaks = seq(0, 100, by = 10),               # 10-year bins
      include.lowest = TRUE,
      right = FALSE,
      labels = paste0(seq(0, 90, 10), "-", seq(10, 100, 10), "y")
    )
  ) %>%
  group_by(flat_type, lease_bin) %>%
  summarise(
    median_psm = median(price_psm, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 30)   # drop sparse tiles

ggplot(hdb_lease_decay, aes(x = lease_bin, y = flat_type, fill = median_psm)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    labels = scales::comma,
    name = "Median Price\n(SGD per sqm)"
  ) +
  labs(
    title = "Lease Decay Heatmap (HDB Resale Prices, 2017–Present)",
    subtitle = "Median price per sqm by flat type and 10-year remaining lease bands",
    x = "Remaining Lease (years, binned)",
    y = "Flat Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )


```
The lease-decay analysis uses the transaction-level HDB resale dataset (2017–2025), as it is the only public dataset containing remaining_lease. Lease decay is a structural relationship and does not require exact alignment with the time ranges of the median-price dataset.


```{r}
# Animated GIF: Evolution of HDB Prices by Town -------------------

library(tidyverse)
library(sf)
library(lubridate)
library(gganimate)
library(gifski)

# Assume you already have:
# df_clean: from your earlier code (MedianResalePricesforRegisteredApplicationsbyTownandFlatType)
# sg_map: planning area shapefile with column `name` in UPPERCASE

# Example (if needed):
# df <- read_csv("datasets/MedianResalePricesforRegisteredApplicationsbyTownandFlatType.csv",
#                na = c("na", "-", "NA"))
# df_clean <- df %>%
#   mutate(
#     price = as.numeric(price),
#     date  = yq(quarter),
#     town  = toupper(town)
#   ) %>%
#   filter(!is.na(price))

# ---- 1. Prepare HDB price data for animation ----

hdb_anim_data <- df_clean %>%
  filter(flat_type == "4-room") %>%       # focus on 4-room; change if needed
  mutate(
    town = toupper(town),
    # optional: restrict to recent years, e.g. last 10–15 years
    year = year(date)
  ) %>%
  # Fix known mismatches between HDB "town" and URA planning areas (sg_map$name)
  mutate(
    town = case_when(
      town == "KALLANG/WHAMPOA" ~ "KALLANG",
      town == "CENTRAL AREA"    ~ "DOWNTOWN CORE",
      TRUE ~ town
    )
  ) %>%
  group_by(date, town) %>%
  summarise(
    median_price = mean(price, na.rm = TRUE),
    .groups = "drop"
  )

# ---- 2. Build full grid of (planning area × date) so "no data" shows as grey ----

all_dates <- sort(unique(hdb_anim_data$date))

full_grid_anim <- expand_grid(
  name = unique(sg_map$name),
  date = all_dates
)

# Join shapes + price data
anim_plot_data <- full_grid_anim %>%
  left_join(sg_map, by = c("name")) %>%
  st_as_sf() %>%
  left_join(hdb_anim_data, by = c("name" = "town", "date"))

# ---- 3. Animated choropleth ----

p_anim <- ggplot(anim_plot_data) +
  geom_sf(aes(fill = median_price), color = "white", size = 0.2) +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    na.value = "grey90",
    labels = scales::comma,
    name = "Median Price (SGD)"
  ) +
  labs(
    title = "Evolution of 4-Room HDB Resale Prices by Town",
    subtitle = "Quarter: {format(frame_time, '%Y-Q%q')}",
    caption = "Grey areas indicate no HDB resale data for that town/quarter"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  ) +
  # animate over time using the 'date' variable
  transition_time(date) +
  ease_aes("linear")

# ---- 4. Render animation (shows in RStudio Viewer) ---- 🔹

anim <- animate(
  p_anim,
  renderer = gifski_renderer(),
  width = 800,
  height = 600,
  fps = 6,
  duration = 12,
  rewind = FALSE
)

anim  # 🔹 printing this makes RStudio display the animation

# ---- 5. Save GIF for use in HTML output ---- 🔹

if (!dir.exists("figures")) dir.create("figures", recursive = TRUE)

anim_save("figures/hdb_4room_prices_by_town.gif", animation = anim)

```
Setup:
```{r}
library(tidyverse)
library(lubridate)
library(stringr)
library(sf)

# 1. Load transaction-level HDB resale data (2017 onwards)
hdb_resale <- read_csv(
  "datasets/ResaleflatpricesbasedonregistrationdatefromJan2017onwards.csv",
  show_col_types = FALSE
)

# 2. Basic cleaning + year, uppercase town
hdb_resale_clean <- hdb_resale %>%
  mutate(
    year = year(ym(month)),
    town = toupper(town)
  )

# 3. Define million-dollar flats
million_flats <- hdb_resale_clean %>%
  filter(resale_price >= 1000000)

```

```{r}
# Aggregate counts of million-dollar flats by town and year
million_by_town_year <- million_flats %>%
  group_by(year, town) %>%
  summarise(n = n(), .groups = "drop") %>%
  # keep towns that have at least 5 million-dollar transactions total (for readability)
  group_by(town) %>%
  mutate(total_n = sum(n)) %>%
  ungroup() %>%
  filter(total_n >= 5)

ggplot(million_by_town_year, aes(x = reorder(town, n), y = n, fill = factor(year))) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Million-Dollar HDB Resale Flats by Town and Year",
    subtitle = "Towns with at least 5 million-dollar transactions (2017–Present)",
    x = "Town",
    y = "Number of Million-Dollar Transactions",
    fill = "Year"
  ) +
  theme_minimal()

```

Figure X shows that million-dollar HDB resale flats are highly concentrated in a small number of towns. Mature, centrally located estates such as [BISHAN / TOA PAYOH / QUEENSTOWN / KALLANG/WHAMPOA — whatever appears in your plot] dominate the count each year, while most towns rarely register any million-dollar transactions. Over time, the number of such transactions generally increases, suggesting that “luxury HDB” units are becoming more common despite repeated cooling measures.


```{r}
# 1. Count million-dollar flats per town (all years combined)
million_by_town <- million_flats %>%
  group_by(town) %>%
  summarise(n = n(), .groups = "drop")

# 2. Fix mismatched names to align with sg_map
million_by_town <- million_by_town %>%
  mutate(
    town_for_map = case_when(
      town == "KALLANG/WHAMPOA" ~ "KALLANG",
      town == "CENTRAL AREA"    ~ "DOWNTOWN CORE",
      TRUE ~ town
    )
  )

# 3. Join with spatial data
sg_map_join <- sg_map %>%
  left_join(million_by_town, by = c("name" = "town_for_map"))

# 4. Compute centroids for point plotting
centroids <- sg_map_join %>%
  st_centroid() %>%
  filter(!is.na(n))   # keep only towns with at least 1 million-dollar flat

# 5. Plot: base map + points sized by count
ggplot() +
  geom_sf(data = sg_map, fill = "grey95", color = "white", linewidth = 0.2) +
  geom_sf(data = sg_map_join, aes(fill = n), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "magma",
    direction = -1,
    na.value = "grey90",
    name = "Million-Dollar\nHDB Resales"
  ) +
  geom_sf(data = centroids, aes(size = n), color = "cyan", alpha = 0.8) +
  scale_size_continuous(name = "Number of\nMillion-Dollar Flats") +
  labs(
    title = "Spatial Concentration of Million-Dollar HDB Flats",
    subtitle = "Counts aggregated over 2017–Present",
    caption = "Circles sized by number of million-dollar transactions; grey areas indicate no such transactions"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "right"
  )

```
Figure Y highlights the spatial clustering of million-dollar HDB flats. Only a handful of towns register a substantial number of million-dollar transactions, and these are predominantly in central or city-fringe regions. Peripheral towns in the North, West, and far East see few or no such sales, reinforcing the idea that location — rather than flat size alone — is a key driver of “luxury” HDB pricing.

```{r}
# For performance, you can optionally downsample non-million points
set.seed(123)
non_million_sample <- hdb_resale_clean %>%
  filter(resale_price < 1000000) %>%
  sample_frac(0.15)   # 15% sample to keep the plot light; adjust as needed

ggplot() +
  geom_point(
    data = non_million_sample,
    aes(x = floor_area_sqm, y = resale_price),
    alpha = 0.15,
    size = 0.6
  ) +
  geom_point(
    data = million_flats,
    aes(x = floor_area_sqm, y = resale_price),
    color = "red",
    alpha = 0.8,
    size = 1.6
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Are Million-Dollar HDB Flats Just Bigger?",
    subtitle = "Floor area vs resale price, highlighting million-dollar transactions",
    x = "Floor Area (sqm)",
    y = "Resale Price (SGD)"
  ) +
  theme_minimal()

```
Figure Z shows the relationship between floor area and resale price. Million-dollar flats (in red) tend to cluster in the upper-right region of the scatterplot, indicating that they are generally larger and more expensive than typical units. However, there is also overlap with non-million-dollar flats of similar size, suggesting that location and flat attributes (storey, orientation, proximity to MRT/amenities) likely push certain units over the million-dollar threshold, rather than size alone.
